---
title: "Study on the Link Between Academic Path and Career Satisfaction"
author: "Ramy Lazghab"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction
This report explores how the alignment between an individual's current role and their skills or education influences career satisfaction. We use Principal Component Analysis (PCA(ACP)) and Multiple Correspondence Analysis (MCA(ACM)) to examine the data.

```{r setup, include=FALSE}
# Load necessary libraries
library(tidyverse)  # Data manipulation
library(FactoMineR) # PCA and MCA functions
library(factoextra) # Visualization for PCA and MCA
library(corrplot)   # Correlation plot
library(dplyr)      # Additional data manipulation functions
```

# Data Loading
We load the dataset from a CSV file and perform an initial inspection to understand its structure.

```{r load-data}
# Load the dataset
data <- read.csv("Data.csv", header = TRUE, sep = ",", fileEncoding = "ISO-8859-1")

# Inspect the structure and content of the dataset
str(data)          # View structure of the dataset
summary(data)      # Get summary statistics
head(data)         # Display the first few rows

# Remove duplicates if any
data <- distinct(data)
```

# Data Preprocessing
We recode Likert-scale responses and Yes/No(oui/Non) columns and prepare subsets of data for PCA(ACP) and MCA(ACM).

## Bloc 1: Pertinence(Pertinence ) (PCA)
This block processes and analyzes the "Pertinence" columns.

```{r pertinence-pca}
# Select Pertinence-related columns
pertinence_columns <- grep("Pertinence", names(data), value = TRUE)
pertinence_data <- data %>%
  select(all_of(pertinence_columns)) %>%
  # Recode Likert responses into numeric values: pdtd -> 1, Neutre -> 2, tafd -> 3
  mutate(across(everything(), 
                ~ recode(., 
                         "pdtd" = 1, 
                         "Neutre" = 2, 
                         "tafd" = 3, 
                         .default = NA_real_)))

# Impute missing values with column means
pertinence_data[is.na(pertinence_data)] <- apply(pertinence_data, 2, function(x) mean(x, na.rm = TRUE))

# Standardize the data to have zero mean and unit variance
pertinence_scaled <- scale(pertinence_data)

# Identify and remove constant/zero-variance columns
col_variances <- apply(pertinence_scaled, 2, var)  # Calculate variances for each column
pertinence_scaled <- pertinence_scaled[, col_variances > 0]  # Remove columns with zero variance

# Perform PCA
pertinence_pca <- prcomp(pertinence_scaled, center = TRUE, scale. = TRUE)

# Scree plot to visualize the proportion of variance explained
fviz_screeplot(pertinence_pca, addlabels = TRUE, ylim = c(0, 50)) +
  ggtitle("Scree Plot: Pertinence")

# Variable biplot: Visualize how variables contribute to the principal components
fviz_pca_var(pertinence_pca, col.var = "cos2") +
  scale_color_gradient2(low = "white", mid = "blue", high = "red", midpoint = 0.6) +
  theme_minimal() +
  ggtitle("Variable Contributions: Pertinence")

# Ensure age is treated as a factor (categorical variable)
data$age_factor <- factor(data$Quel.est.votre.age.., 
                          levels = c("18-24", "25-34", "35-44", "45-54", "55+", "Autre"))

# Filter the 'age' column to include only complete cases in pertinence_data
age_filtered <- data$age_factor[complete.cases(pertinence_data)]

# Ensure the length of age_filtered matches the number of individuals in pertinence_scaled
age_filtered <- age_filtered[1:nrow(pertinence_scaled)]

# Create the PCA plot colored by age with a discrete color palette
fviz_pca_ind(pertinence_pca, geom = "point", habillage = age_filtered) +
  scale_color_brewer(palette = "Set3") +
  ggtitle("Pertinence PCA: Individuals Colored by Age") + 
  theme_minimal()

# Create the PCA plot colored by age with custom colors
fviz_pca_ind(pertinence_pca, geom = "point", habillage = age_filtered) +
  scale_color_manual(values = c("18-24" = "blue", "25-34" = "green", "35-44" = "red", 
                               "45-54" = "purple", "55+" = "orange", "Autre" = "gray")) +
  ggtitle("Pertinence PCA: Individuals Colored by Age") + 
  theme_minimal()

# Individual biplot (colored by education level)
fviz_pca_ind(pertinence_pca, geom = "point", habillage = data$Quel.est.votre.niveau.d.etude) +
  scale_color_brewer(palette = "Set1") +
  ggtitle("Pertinence PCA: Individuals Colored by Education Level") + theme_minimal()
```

## Bloc 2: Utilisation(Utilisation) (PCA)
This block processes and analyzes the "Utilisation" columns.

```{r utilisation-pca}
# Select Utilisation-related columns
utilisation_columns <- grep("Utilisation", names(data), value = TRUE)
utilisation_data <- data %>%
  select(all_of(utilisation_columns)) %>%
  # Recode Likert responses into numeric values: pdtd -> 1, Neutre -> 2, tafd -> 3
  mutate(across(everything(), 
                ~ recode(., 
                         "pdtd" = 1, 
                         "Neutre" = 2, 
                         "tafd" = 3, 
                         .default = NA_real_)))

# Impute missing values with column means
utilisation_data[is.na(utilisation_data)] <- apply(utilisation_data, 2, function(x) mean(x, na.rm = TRUE))

# Standardize the data to have zero mean and unit variance
utilisation_scaled <- scale(utilisation_data)

# Identify and remove constant/zero-variance columns
col_variances <- apply(utilisation_scaled, 2, var)  # Calculate variances for each column
utilisation_scaled <- utilisation_scaled[, col_variances > 0]  # Remove columns with zero variance

# Perform PCA
utilisation_pca <- prcomp(utilisation_scaled, center = TRUE, scale. = TRUE)

# Scree plot to visualize the proportion of variance explained
fviz_screeplot(utilisation_pca, addlabels = TRUE, ylim = c(0, 50)) +
  ggtitle("Scree Plot: Utilisation")

# Variable biplot: Visualize how variables contribute to the principal components
fviz_pca_var(utilisation_pca, col.var = "cos2") +
  scale_color_gradient2(low = "white", mid = "blue", high = "red", midpoint = 0.6) +
  theme_minimal() +
  ggtitle("Variable Contributions: Utilisation")

# Individual biplot (colored by age)
fviz_pca_ind(utilisation_pca, geom = "point", habillage = data$age) +
  scale_color_brewer(palette = "Set3") +
  ggtitle("Utilisation PCA: Individuals Colored by Age") + theme_minimal()

# Individual biplot (colored by education level)
fviz_pca_ind(utilisation_pca, geom = "point", habillage = data$Quel.est.votre.niveau.d.etude) +
  scale_color_brewer(palette = "Set1") +
  ggtitle("Utilisation PCA: Individuals Colored by Education Level") + theme_minimal()

```

## Bloc 3: Necessite(Necessity) (PCA)
This block processes and analyzes the "Necessite" columns.

```{r necessite-pca}
# Select Necessite-related columns
necessite_columns <- grep("Necessite", names(data), value = TRUE)
necessite_data <- data %>%
  select(all_of(necessite_columns)) %>%
  # Recode Likert responses into numeric values: pdtd -> 1, Neutre -> 2, tafd -> 3
  mutate(across(everything(), 
                ~ recode(., 
                         "pdtd" = 1, 
                         "Neutre" = 2, 
                         "tafd" = 3, 
                         .default = NA_real_)))

# Impute missing values with column means
necessite_data[is.na(necessite_data)] <- apply(necessite_data, 2, function(x) mean(x, na.rm = TRUE))

# Standardize the data to have zero mean and unit variance
necessite_scaled <- scale(necessite_data)

# Identify and remove constant/zero-variance columns
col_variances <- apply(necessite_scaled, 2, var)  # Calculate variances for each column
necessite_scaled <- necessite_scaled[, col_variances > 0]  # Remove columns with zero variance

# Perform PCA
necessite_pca <- prcomp(necessite_scaled, center = TRUE, scale. = TRUE)

# Scree plot to visualize the proportion of variance explained
fviz_screeplot(necessite_pca, addlabels = TRUE, ylim = c(0, 50)) +
  ggtitle("Scree Plot: Necessite")

# Variable biplot: Visualize how variables contribute to the principal components
fviz_pca_var(necessite_pca, col.var = "cos2") +
  scale_color_gradient2(low = "white", mid = "blue", high = "red", midpoint = 0.6) +
  theme_minimal() +
  ggtitle("Variable Contributions: Necessite")

# Individual biplot (colored by education level)
fviz_pca_ind(necessite_pca, geom = "point", habillage = data$Quel.est.votre.niveau.d.etude) +
  scale_color_brewer(palette = "Set1") +
  ggtitle("Necessite PCA: Individuals Colored by Education Level") + theme_minimal()

```
# Select Oui/Non columns for MCA(ACM in FR)

```{r}
objectifs_columns <- grep("Objectifs", names(data), value = TRUE)
objectifs_data <- data %>%
  select(all_of(objectifs_columns)) %>%
  # Convert to factors for MCA
  mutate(across(everything(), as.factor))
```

# Remove columns with constant values or all NAs
```{r}
objectifs_data <- objectifs_data[, sapply(objectifs_data, function(col) length(unique(na.omit(col))) > 1)]

```

# Perform MCA
```{r}
objectifs_mca <- MCA(objectifs_data, graph = FALSE)
```


# Scree plot for MCA
```{r}
fviz_screeplot(objectifs_mca, addlabels = TRUE, ylim = c(0, 50)) +
  ggtitle("Scree Plot: Objectifs MCA")
```


# MCA biplot: Variables and individuals
```{r}
fviz_mca_biplot(objectifs_mca, geom = "point", col.var = "blue", col.ind = "red") +
  ggtitle("MCA Biplot: Objectifs") + theme_minimal()
```


# MCA biplot (colored by age)
```{r}
fviz_mca_ind(objectifs_mca, geom = "point", habillage = data$age, palette = "Set3") +
  ggtitle("Objectifs MCA: Individuals Colored by Age") + theme_minimal()
```


# MCA biplot (colored by education level)
```{r}
# Ensure the education level column is a factor
data$Quel.est.votre.niveau.d.etude <- as.factor(data$Quel.est.votre.niveau.d.etude)

# Create the MCA biplot colored by education level
fviz_mca_ind(objectifs_mca, geom = "point", habillage = data$Quel.est.votre.niveau.d.etude, palette = "Set1") +
  ggtitle("Objectifs MCA: Individuals Colored by Education Level") + theme_minimal()

```
#Data Preparation for AFC
I select appropriate categorical variables for correspondence analysis.
to see the study level and the Necessity for Learning new skills
```{r}
# Correct column names
categorical_columns <- grep("Quel.est.votre.niveau.d.etudes..|Necessite.d.un.Apprentissage.ou.d.une.Adaptation....J.ai.du.apprendre.de.nouvelles.compétences.pour.repondre.aux.exigences.de.mon.poste..", colnames(data), value = TRUE)


# Select the columns and convert to factors
afc_data <- data %>%
  select(all_of(categorical_columns)) %>%
  mutate(across(everything(), as.factor))  # Ensure variables are factors

# Create the contingency table for AFC
afc_table <- table(afc_data[[1]], afc_data[[2]])

# Perform Correspondence Analysis (AFC)

afc_result <- CA(afc_table)
# Plot the Correspondence Analysis
fviz_ca_biplot(afc_result, 
               col.row = "blue",    # Color rows (study level)
               col.col = "red",     # Color columns (Necessity for learning new skills)
               title = "Correspondence Analysis (AFC)")

# Display results
summary(afc_result)

```
#Conclusion
This analysis demonstrates how the application of PCA and the exploration of the dataset help us understand the relationship between the academic background, role satisfaction, and other variables.

